---
title: "ANOVA"
---

*Last modified on `r Sys.Date()`*

```{r, echo = FALSE, message = FALSE}
source("libs/Common.R")
```

```{r Load fonts, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
library(extrafont)
font_import(prompt=FALSE, pattern="[A/a]rchitects")
loadfonts(device="postscript")

```

----

# Introduction
Analysis of Variance (ANOVA) seeks to compare the means between two or more batches of numbers. You can think of an ANOVA as an extension of the t-test where three or more batches need to be compared. The name may seem misleading since it suggests that we are comparing variances and not some central value, but in fact, we compare the variances (spreads) between batches to assess if the central values are significantly different from one another. For example, let's compare the following batches of numbers:


```{r, echo=FALSE, fig.height=2.5, fig.width=5}
set.seed(3)
x <- sample(c(23.5,24,24.2,25,25.3))
y <- sample(c(24.5, 25.2, 25.8, 26.0, 26.5))
z <- sample(c(23.1,23.6,24.2,24.6,25))

x.x <- seq(0.5,1.3,.2)
y.x <- seq(2,2.8,.2)
z.x <- seq(3.2,4.0,.2)

mean.grand <- mean(c(x,y,z))

OP <- par(mar=c(2,3,1,4))
plot(c(x,y,z)~ jitter(c(rep(1,length(x)),rep(2,length(y)),rep(3,length(z))),0.1),
     pch=215, ylab=NA,xlab=NA,axes = F,family="Architects Daughter", cex=1.3,
     xlim=c(0,3.5), ylim=c(23,27))
axis(1, at=c(1,2,3),label=c("Batch 1", "Batch 2", "Batch 3"),col="grey", 
     family="Architects Daughter" ,tick=FALSE, cex.axis=0.8)
axis(2, at=c(23,24,25,26,27), family="Architects Daughter", las=2, 
     col="grey80", col.axis="grey80", cex.axis=0.8)
par(OP)
```

Given the very small overlap in spread between batch 2 and the two other batches, it's obvious that batch 2 is significantly different from batches 1 and 3, but can we say with a similar level of certainty that batch 1 is significantly different from batch 3? Probably not, the slight offset in spread between batch 1 and 3 may be due to chance alone.

A one-way ANOVA compares measurement means between a single group of levels or batches. ANOVAs can be extended to include multiple groups (each having different levels). An ANOVA that compares means between two groups (each having their own set of levels) is referred to a two-way ANOVA. 


# How a one-way ANOVA is calculated

This section focuses on one group of levels (hence a one-way ANOVA).

## The variance-ratio method

An ANOVA test seeks to compare the spread between the batches (technically referred to as *levels*). 

The first step is to sum the square of the *distances* between each value (from all levels) to the *grand* mean computed from *all* values (plotted as a dark dashed line in the following graphic). We'll call this value the *total sum of squares* for the mean ($SSE_{mean}$). It's calculated as follows:


$$
SSE_{mean} = \sum (y - \bar{\bar y})^2 
$$

```{r, echo=FALSE}
mean.grand <- mean( c(x,y,z) )
SSE.m <- sum( (c(x,y,z) - mean.grand)^2 ) # Total sum of squares
```

where $\bar{\bar{y}}$ is the mean for *all* values. In this example, $\bar{\bar y}$ equals `r round(mean.grand,1)`. In the following plot, we spread out the values in each level (for clarity) and measure their distances to the grand mean. Each level is assigned a unique color to distinguish the different batches.

```{r, echo=FALSE, fig.height=3, fig.width=5}
OP <- par(mar=c(2,3,1,4))
plot(c(x,y,z)~ jitter(c(x.x,y.x,z.x),0.1),
     pch=215, ylab=NA,xlab=NA,axes = F,family="Architects Daughter", cex=1.3,
     xlim=c(0,5.5), ylim=c(23,27))
axis(1, at=c(0.8,2.3,3.75),label=c("Batch 1", "Batch 2", "Batch 3"), 
     family="Architects Daughter", tick=FALSE , cex.axis=.8)
axis(2, at=c(23,24,25,26,27), family="Architects Daughter", las=2, 
     col="grey80", col.axis="grey80", cex.axis=.8)
abline(h=mean(c(x,y,z)),lty=2)
for (i in 1:length(x)){
  lines(rep(x.x[i],2), c(x[i],mean.grand), col="red")
}
for (i in 1:length(y)){
  lines(rep(y.x[i],2), c(y[i],mean.grand), col="green")
}
for (i in 1:length(z)){
  lines(rep(z.x[i],2), c(z[i],mean.grand), col="blue")
}
text(4.1,26.5, "SSE-mean\n(Difference between\nall values\nand grand mean)",
     family="Architects Daughter", cex=0.8)
text(4.8, mean.grand+.15, mean.grand, family="Architects Daughter", cex=0.8)

par(OP)
```

Next, we compare the values in each level to their respective level means. Similar to the way we compute the $SSE_{mean}$, we sum the squared differences for each level as follows:

$$
SSE = \sum (y_{batch1} - \bar y_{batch1})^2 + \sum (y_{batch2} - \bar y_{batch2})^2 + \sum (y_{batch3} - \bar y_{batch3})^2
$$

Where $SSE$ is the **error sum of squares**.

```{r, echo=FALSE, fig.height=3, fig.width=5}
b1.mean <- mean(x)
b2.mean <- mean(y)
b3.mean <- mean(z)

OP <- par(mar=c(2,3,1,4))
plot(c(x,y,z)~ jitter(c(x.x,y.x,z.x),0.1),
     pch=215, ylab=NA,xlab=NA,axes = F,family="Architects Daughter", cex=1.3,
     xlim=c(0,5.5), ylim=c(23,27))
axis(1, at=c(0.8,2.3,3.75),label=c("Batch 1", "Batch 2", "Batch 3"), 
     family="Architects Daughter", tick=FALSE , cex.axis=.8)
axis(2, at=c(23,24,25,26,27), family="Architects Daughter", las=2, 
     col="grey80", col.axis="grey80", cex.axis=.8)

abline(h=mean(c(x,y,z)),lty=2, col="grey40")

lines(c(x.x[1],x.x[5]), c(b1.mean,b1.mean), col="red")
lines(c(y.x[1],y.x[5]), c(b2.mean,b2.mean), col="green")
lines(c(z.x[1],z.x[5]), c(b3.mean,b3.mean), col="blue")

for (i in 1:length(x)){
  lines(rep(x.x[i],2), c(x[i],b1.mean), col="red")
}
for (i in 1:length(y)){
  lines(rep(y.x[i],2), c(y[i],b2.mean), col="green")
}
for (i in 1:length(z)){
  lines(rep(z.x[i],2), c(z[i],b3.mean), col="blue")
}
text(4.0,26.5, "SSE\n(Difference between\nbatch values\nand batch means)",
     family="Architects Daughter", cex=0.8)
text(4.8, mean.grand+.15, mean.grand, family="Architects Daughter", cex=0.8)

text(x.x[1]-.35, b1.mean, b1.mean, family="Architects Daughter", col="red", cex=.8)
text(y.x[1]-.35, b2.mean, b2.mean, family="Architects Daughter", col="green", cex=.8)
text(z.x[1]-.35, b3.mean, b3.mean, family="Architects Daughter", col="blue", cex=.8)


par(OP)
```

If the mean values of all three levels are the same, their horizontal lines should line up with the grand mean and both $SSE_{mean}$ and $SSE$ should be equal, if not, $SSE$ will be less than $SSE_{mean}$ (the distance between the points and their respective level mean will always be equal or shorter than their distances to the overall mean). The difference between $SSE_{mean}$ and $SSE$ is called the **treatment sum of squares** (also referred to as the **model sum of squares**):

$$
SSR = SSE_{mean} - SSE
$$

If SSR is close to 0, then the differences between the levels is small, if SSR is large, then two or more of the levels are significantly different from one another. SSR is, in fact, the squared difference between each group's mean and the grand mean. Think of two competing models to predict the observed values: the grand mean, and each level's mean. If the level means are close to the grand mean, then both the level means and grand mean will generate the same expected values. If they are different, then they will generate different expected values. So the greater the SSR, the more different the predicted values (and therefore the more different the means).

```{r echo=FALSE, fig.height=3, fig.width=5}
OP <- par(mar=c(2,3,1,4))
plot(c(x,y,z)~ jitter(c(x.x,y.x,z.x),0.1),
     pch=215, ylab=NA,xlab=NA,axes = F,family="Architects Daughter", cex=1.3,
     xlim=c(0,5.5), ylim=c(23,27))
axis(1, at=c(0.8,2.3,3.75),label=c("Batch 1", "Batch 2", "Batch 3"), 
     family="Architects Daughter", tick=FALSE , cex.axis=.8)
axis(2, at=c(23,24,25,26,27), family="Architects Daughter", las=2, 
     col="grey80", col.axis="grey80", cex.axis=.8)

abline(h=mean(c(x,y,z)),lty=2, col="grey40")

lines(c(x.x[1],x.x[5]), c(b1.mean,b1.mean), col="red")
lines(c(y.x[1],y.x[5]), c(b2.mean,b2.mean), col="green")
lines(c(z.x[1],z.x[5]), c(b3.mean,b3.mean), col="blue")

for (i in 1:length(x)){
  lines(rep(x.x[i],2), c(mean.grand,b1.mean), col="red")
}
for (i in 1:length(y)){
  lines(rep(y.x[i],2), c(mean.grand,b2.mean), col="green")
}
for (i in 1:length(z)){
  lines(rep(z.x[i],2), c(mean.grand,b3.mean), col="blue")
}
text(4.0,26.5, "SSR\n(Difference between\nmodeled values\nand grand mean)",
     family="Architects Daughter", cex=0.8)
text(4.8, mean.grand+.15, mean.grand, family="Architects Daughter", cex=.8)

text(x.x[1]-.35, b1.mean, b1.mean, family="Architects Daughter", col="red", cex=.8)
text(y.x[1]-.35, b2.mean, b2.mean, family="Architects Daughter", col="green", cex=.8)
text(z.x[1]-.35, b3.mean, b3.mean, family="Architects Daughter", col="blue", cex=.8)
par(OP)

```

The goal is to compare the variability in $SSE$ to that in $SSR$, however, $SSR$ can take on much larger values than $SSA$ (because $SSR$ is measuring from the group means and not the individual values). To remedy this bias, we compute the **mean squares** from both values by dividing the sum of squares by the **degrees of freedom**. For $SSR$, this gives us:
$$
MSR = \frac{SSR}{p-1}
$$

where $MSR$ is the **mean square for treatments** (or **mean square for model**), and $p$ is the number of levels (3 in this example). 

Likewise, we can compute the **mean square for error** as:

$$
MSE = \frac{SSE}{n-p}
$$

where $n$ is the total number of observations.

Next, we compute the $F$-statistic (or $F$-ratio) as:

$$
F = \frac{MSR}{MSE}
$$

$F$ gives us the proportion of the overall spread in the measurements that can be explained by the levels vs. the proportion of the overall spread in the measurements *not* explained by the levels. A value of $F$ that approaches $1$ indicates that little to none of the variability in the measurements can be explained by the levels  suggesting that differences in their mean may be due to random noise only. If $F$ is much larger than one, then the spreads between levels are quite different (meaning that these differences are large contributors to the overall spread) suggesting that the observed differences in mean values are significant too.

The following graphic shows the spread for *all* values in the data (right-side plot) and its spread broken down by levels (left-side plot). In this case, it seems that a good chunk of the overall spread can be explained by the differences in spread between levels. If this is still unclear, picture the plots with the measurements for batch 2 removed; what you would note is that the overall spread will be noticeably reduced suggesting that the measurements in level two alone were an important contributor to the overall spread.

```{r fig.width=9, fig.height=2.5, fig.show='hold', echo=FALSE}
set.seed(4)
OP <- par(mar=c(2,2,1,1), mfrow=c(1,2))
plot(c(x,y,z)~ jitter(c(rep(1,length(x)),rep(2,length(y)),rep(3,length(z))),0.1),
     pch=215, ylab=NA,xlab=NA,axes = F,family="Architects Daughter", cex=1.3,
     xlim=c(0,3.5), col=c(rep("red",5),rep("green",5),rep("blue",5)), ylim=c(23,27))
axis(1, at=c(1,2,3),label=c("Batch 1", "Batch 2", "Batch 3"), 
     family="Architects Daughter", tick=FALSE )
axis(2, at=c(23,24,25,26,27), family="Architects Daughter", las=2, 
     col="grey80", col.axis="grey80", cex.axis=.8)

plot(c(x,y,z) ~ jitter(rep(1,length(c(x,y,z))), 2),
     pch=215, ylab=NA,xlab=NA,axes = F,family="Architects Daughter", cex=1.3,
     xlim=c(0,3.5), col=c(rep("red",5),rep("green",5),rep("blue",5)), ylim=c(23,27))
axis(1, at=c(1),label=c("All values"), family="Architects Daughter", tick=FALSE )
text(2,25,"F >> 1", family="Architects Daughter")
par(OP)


```

Contrast this last example with the following dataset where the differences in spread (and mean) between levels are negligible. Removing anyone of the batches from the dataset would have a negligible impact on the overall spread in the measurements. We would say that little of the spread (variability) in the overall measurements can be explained by differences in measurements between levels; this would result in an $F$-ratio close to $1$.

```{r fig.width=9, fig.height=2.5, fig.show='hold', echo=FALSE}
set.seed(4)
x1 <- x
y1 <- y-1.1
z1 <- z

OP <- par(mar=c(2,2,1,1), mfrow=c(1,2))
plot(c(x1,y1,z1)~ jitter(c(rep(1,length(x1)),rep(2,length(y1)),rep(3,length(z1))),0.1),
     pch=215, ylab=NA,xlab=NA,axes = F,family="Architects Daughter", cex=1.3,
     xlim=c(0,3.5), col=c(rep("red",5),rep("green",5),rep("blue",5)), ylim=c(23,26))
axis(1, at=c(1,2,3),label=c("Batch 1", "Batch 2", "Batch 3"), 
     family="Architects Daughter", tick=FALSE)
axis(2, at=c(23,24,25,26), family="Architects Daughter", las=2, 
     col="grey80", col.axis="grey80", cex.axis=.8)

plot(c(x1,y1,z1) ~ jitter(rep(1,length(c(x1,y1,z1))),2),
     pch=215, ylab=NA,xlab=NA,axes = F,family="Architects Daughter", cex=1.3,
     xlim=c(0,3.5), col=c(rep("red",5),rep("green",5),rep("blue",5)), ylim=c(23,26))
axis(1, at=c(1),label=c("All values"), family="Architects Daughter", tick=FALSE )
text(2,25,"F = 1", family="Architects Daughter")
par(OP)
```

To assess whether a value of $F$ is significantly greater than 1, we must compare our observed $F$ to a distribution of $F$'s we would expect to get if the means between levels were *not* different (we refer to the hypothesized $F$ values as $F_{Ho}$). 

In the following plot, a hypothetical $F$ ratio (plotted as a vertical red line) is located to the right of the distribution of $F_{Ho}$ values (delineated in a black line in the following plot). The shaded pink area to the right of the hypothetical $F$ represents the fraction of $F_{Ho}$ that would be more extreme than our observed $F$. The goal is to assess whether or not we feel comfortable in concluding that our observed $F$ is significantly different from an $F$ we would expect to get if the means between batches were all the same.

```{r echo=FALSE, message=FALSE, fig.width=7, fig.height=3.0, warning=FALSE}
OP <- par("mar"=c(2,4,3,1), xpd=NA)
df1 = 7
df2 = 5
Ft  = 2.12  
xmin = 0
xmax = 10
p10  = qf(0.10,df1,df2)
p025  = qf(0.025,df1,df2)
p975  = qf(0.975,df1,df2)
ncurve.x = seq(xmin, xmax,length.out= 200)
ncurve.y = df(ncurve.x, df1, df2 )
plot( ncurve.y ~ ncurve.x, type="l", ylab=NA, xlab=NA,axes=FALSE,main=NA)
axis(1, family= "Architects Daughter")
lines(x = c(Ft,Ft), y=c(0,df(Ft,df1,df2)), col="red"  , lty=1, lw=2)
lines(x = c(p975,p975) , y=c(0,df( p975,df1,df2)) , col="grey", lty=2, lw=2)

text(x=Ft+.7, y = df(Ft,df1,df2)+.3, "Distribution of F(Ho)", col="black", pos=3, family= "Architects Daughter", adj=10.5)
text(x=Ft+1.3, y = df(Ft,df1,df2), "Observed F-value", col="red", pos=3, family= "Architects Daughter", adj=10.5)

ncurve.x.mSE <- seq(Ft, xmax, length.out= 100)
ncurve.y.mSE <- df(ncurve.x.mSE, df1, df2 )
ncurve.x.mSE <- c(ncurve.x.mSE, max(ncurve.x.mSE), min(ncurve.x.mSE))
ncurve.y.mSE <- c(ncurve.y.mSE, 0,0)
polygon( ncurve.x.mSE, ncurve.y.mSE, col=rgb(1, 0, 0,.3), border=NA)
par(OP)
```

## An example in R

In this working example, we compare fecal coliform counts (represented as the log10 of organisms per 100 ml) in the Illinois River between seasons (summer, fall, winter and spring) across six years (data from Millard and Neerchal, 2001).


Year |   Summer |   Fall |  Winter |  Spring
----- | -------- | ------ | -------- | -------
1971 |   2.00 | 1.45  | 1.45  | 1.34
1972 | 2.34 | 2.08  | 1.76  | 1.72
1973 | 2.48 | 2.32  | 2.08  | 2.04
1974 | 2.63 | 2.45  | 2.36  | 2.15
1975 |  2.81 | 2.70 |  2.49 |  2.51
1976 | 3.20 | 3.04  | 2.70  | 3.11

Let's first generate the data. We will create two data frames of the same data. One will be in a wide format, the other in a long format. We will use the former to generate a plot and the latter will be used in the ANOVA analysis.

```{r, tidy=FALSE}
# Create a data frame
dat <- data.frame( Year = c(1971,1972,1973,1974,1975,1976),
                   Summer = c(2.00, 2.34, 2.48, 2.63, 2.81, 3.20),
                   Fall = c(1.45, 2.08, 2.32, 2.45, 2.70, 3.04),
                   Winter = c(1.45, 1.76, 2.08, 2.36, 2.49, 2.70),
                   Spring = c(1.34, 1.72, 2.04, 2.15, 2.51, 3.11))

# However, to run an ANOVA, the data needs to be in a long form
library(tidyr)
dat.long <- gather(dat, key = "Season", value="Value", -Year)
```

Next, let's plot the points then add the grand mean (black line) and each season's mean (red lines) to the plot.

```{r plot.example1, fig.width=4, fig.height=3, eval=FALSE}

# Create an empty plot
plot( NULL, xlim= c(0.5,4.5), ylim=c(1, 3.5), axes=FALSE, xlab=NA,ylab="Fecal Coliform")
box()
axis(1, labels = names(dat)[-1], at=c(1,2,3,4))
axis(2, las=2, )
points(rep(1,nrow(dat)), dat$Summer, pch=1)
points(rep(2,nrow(dat)), dat$Fall, pch=2)
points(rep(3,nrow(dat)), dat$Winter, pch=3)
points(rep(4,nrow(dat)), dat$Spring, pch=4)

# Add the grand mean
abline(h = mean(dat.long$Value), lty=3)

# Add each group's mean
lines(c(.8,1.2), rep(mean(dat$Summer),2), col="red" )
lines(c(1.8,2.2), rep(mean(dat$Fall),2), col="red" )
lines(c(2.8,3.2), rep(mean(dat$Winter),2), col="red" )
lines(c(3.8,4.2), rep(mean(dat$Spring),2), col="red" )
```


```{r fig.width=4, fig.height=3, eval=TRUE, echo=FALSE}
OP <-par(mar=c(2,3,1,1))
 <<plot.example1>>
par(OP)
```

A quick glance at the plot suggests that the variance within each season is large relative to the differences in means between each seasons. But could summer values be slightly higher than those of other seasons? We check this assumption with an ANOVA test.

### Computing ANOVA the hard way

```{r, tidy=FALSE}
SSE.m <- sum( (dat.long$Value - mean(dat.long$Value))^2 ) 
SSE <- sum( (dat$Summer - mean(dat$Summer))^2 +
            (dat$Fall - mean(dat$Fall))^2 +
            (dat$Winter - mean(dat$Winter))^2 + 
            (dat$Spring - mean(dat$Spring))^2 )  
SSR <- SSE.m - SSE
MSR <- SSR / 3   # (p - 1) or 3 degrees of freedom
MSE <- SSE / 20  # (n - p) or 20 degrees of freedom
Fratio <- MSR/MSE
p.val  <- pf(Fratio, 3, 20,lower.tail=FALSE)
```

The function `pf()` compares our $F$-ratio to the distribution of $F_{Ho}$ values one would expect to get if the means between seasons were not different. The two numbers following the $F$-ratio value are the degrees of freedom for the $MSR$ and the $MSE$ calculated from $(p-1)$ and $(n-p)$ respectively.


```{r echo=FALSE, message=FALSE, fig.width=7, fig.height=2.0, warning=FALSE}
OP <- par("mar"=c(2,4,1,1), xpd=NA)
df1 = 3
df2 = 20
Ft  = Fratio  
xmin = 0
xmax = 10
ncurve.x = seq(xmin, xmax,length.out= 200)
ncurve.y = df(ncurve.x, df1, df2 )
plot( ncurve.y ~ ncurve.x, type="l", ylab=NA, xlab=NA,axes=FALSE,main=NA)
axis(1, family= "Architects Daughter")
lines(x = c(Ft,Ft), y=c(0,df(Ft,df1,df2)), col="red"  , lty=1, lw=2)
lines(x = c(p975,p975) , y=c(0,df( p975,df1,df2)) , col="grey", lty=2, lw=2)


text(x=Ft+1.2, y = df(Ft,df1,df2), "Our F-value", col="red", pos=3, family= "Architects Daughter", adj=10.5)

ncurve.x.mSE <- seq(Ft, xmax, length.out= 100)
ncurve.y.mSE <- df(ncurve.x.mSE, df1, df2 )
ncurve.x.mSE <- c(ncurve.x.mSE, max(ncurve.x.mSE), min(ncurve.x.mSE))
ncurve.y.mSE <- c(ncurve.y.mSE, 0,0)
polygon( ncurve.x.mSE, ncurve.y.mSE, col=rgb(1, 0, 0,.3), border=NA)
par(OP)
```

The computed statistics are:

Statistic   Value
---------   -----
SSE(mean)   `r round(SSE.m,2)`
SSE         `r round(SSE,2)`
SSR         `r round(SSR,2)`
MSR         `r round(MSR,2)`
MSE         `r round(MSE,2)`
F           `r round(Fratio,2)`
p           `r round(p.val,3)`

Our calculated p-value is `r round(p.val,3)` indicating that about `r round(p.val * 100)`% of the $F_{Ho}$ values are more extreme than ours. So it would be unwise to dismiss the chance that the means between all four seasons are equal to one another.

### Computing ANOVA the easy way

We can use the `anova` function to compute the $F$-ratio and the $p$-value. The function takes as argument a model (a linear regression model in this case) where the dependent variable $y$ is the measurement value and the independent variable $x$ is the level (or seasons in our example). This implementation of ANOVA requires that the season values be in one column (the $x$ column) and that the measurements be in another column (the $y$ column). This requires that we use the *long* version of our table, `dat.long`, where the $x$ column is labeled `Season` and the $y$ column is labeled `Value`. The first few rows of `dat.long` look like this:
```{r}
head(dat.long)
```

The ANOVA analysis is thus computed as:

```{r}
anova(lm(Value ~ Season, dat.long))
```

The column `Mean Sq` displays the mean sum-of-squares for treatment  $SSR$, and the error sum-of-square, $SSE$. The $F$-ratio and $p$-value are the same as those computed in the last subsection. Again, there is no evidence that the seasons have an influence on the mean concentrations of fecal coliform counts.

### ANOVA as a regression

You'll note that this approach in computing the ANOVA makes use of the linear regression function `lm`. This is because a one-way ANOVA is nothing more than a regression between all values in the batches and their levels expressed as categorical values where the number of categorical values is the number of levels minus $1$. In essence the ANOVA is generating the following model:

$$
Coliform\ count = a + b(FALL) + c(Winter) + d(Spring)
$$

So, if a value belongs to the FALL batch, the model looks like this:

$$
Coliform\ count = a + b(1) + c(0) + d(0)
$$

If the value belongs to the SUMMER batch, the model looks like this:

$$
Coliform\ count = a + b(0) + c(0) + d(0)
$$

where $a$ is the mean value for summer coliform measurements. It follows that the coefficients $b$, $c$ and $d$ are differences in mean values between summer measurements and fall, winter and spring measurements respectively:

$a$ = `mean(dat$Summer)` = `r round(mean(dat$Summer),2)`

$b$ = `mean(dat$Fall) - mean(dat$Summer)` = `r round(mean(dat$Fall)- mean(dat$Summer),2)`

$c$ = `mean(dat$Winter) - mean(dat$Summer)` = `r round(mean(dat$Winter)- mean(dat$Summer),2)`

$d$ = `mean(dat$Spring) - mean(dat$Summer)` = `r round(mean(dat$Spring)- mean(dat$Summer),2)`

These coefficients can be extracted from the `lm` model via the `coefficients` function.

```{r, tidy=FALSE}
coefficients( lm( Value ~ Season, dat.long) )
```

# Assumptions of ANOVA

There are three assumptions that should be met when computing an ANOVA:

1. The variance between the batches (homogeneity of variance) should be similar.
2. Observations should be independent of one another.
3. The distribution of values within each group should be normally distributed. If a transformation is applied to the data, it should be applied to *all* batches. Note that the distribution of values for all combined batches need not be normally distributed.

# Identifying which levels are different

So far, we have seen that the ANOVA test tells us if the means between batches (levels) are similar, but it does not tell us *which* batch(es) of numbers may be different. One approach in assessing which level(s) is(are) significantly different from the mean is to implement a **post hoc** procedure (aka **pairwise comparison**).

There are many *post hoc* procedures to choose from. Two popular methods are presented here: the **Bonferroni** procedure and **Tukey's HSD** procedure. Both tend to generate slightly different results depending on the data characteristics. It's therefore worthwhile to run both to assess if the levels are significantly different from one another. But note that in many cases there is no substitute for a visual assessment of the differences between batches.

## Bonferroni

The Bonferroni method can be implemented using the `pairwise.t.test` function with the first two parameters pointing to the column of values and the column of variables respectively, and the third parameter pointing to the pairwise method.

```{r tidy=FALSE}
pairwise.t.test( dat.long$Value, dat.long$Season, p.adjust.method = "bonferroni")
```

The output is a matrix of $p$-values for different pairs of levels (Fall-Summer, Winter-Summer, etc...). A low $P$-value indicates significant differences between levels. In our working example, all $P$-values are relatively high indicating that the means between levels are not statistically significant.

## Tukey's HSD

The Tukey method is implemented using the `TukeyHSD` method. The first parameter is an `aov` object (`aov` is nearly identical to the `anova(lm(...))` method used thus far), and the second variable is the *levels* column from the data.

```{r tidy=FALSE}
TukeyHSD( aov( Value ~ Season, dat.long), "Season")
```

The Tukey method generates a table of $P$-values (as opposed to a matrix) for all pairs of levels. Note the slightly lower $P$-values than those generated with the Bonferroni method. The Tukey HSD method tends to be more conservative than the Bonferri method and is thus more likely to reject the null hypothesis (that the means are equal) than the Bonferri method. This comes with an advantage, however, in that the Tukey test tends to have greater statistical *power* when there are many different levels (or batches) in our dataset.

The Tukey test also generates the 95% confidence intervals (`lwr` = lower bound and `upr` = upper bound). If the lower bound is less than $0$ and the upper bound is greater than $0$ we cannot say that the means between both levels are not significantly different at a confidence level of 0.05.

## Assumptions of the post hoc test

The *post* hoc procedure assumes that we are only interested in knowing which effect (if any) is different from the grand mean. It makes no assumption about the direction of this difference (i.e. if the level mean is greater than or less than the overall mean). This is analogous to implementing a two-tailed test. 

If knowing whether the effect mean is greater than or less than the overall mean, a **planned contrast** procedure (aka **planned comparison** procedure)  should be adopted instead. This procedure is not covered in this tutorial, but can be found in most introductory stats books. 

# Another working example

The coliform dataset was an example of effects that did not differ. Here, we'll look at an example with differences in effect means. We'll use a dataset that comes installed with `R`, the `ChickWeight` data. This dataset tracks the weights of chicks using different diet types. We will grab a subset of the data and only focus on day 21 measurements. We will also restrict the columns to `Weight` and `Diet`.

```{r}
dat <- ChickWeight[ChickWeight$Time==21,c("weight", "Diet")]
```

The group column is `Diet` which has four levels. These levels are identified as numbers ($1$ through $4$). One needs to be careful when running an ANOVA using levels encoded as numbers since the `lm` model may interpret such values as continuous and not categorical. It's therefore good practice to force the numbers as `factors` by using the `as.factor` function:

```{r}
dat$Diet <- as.factor(dat$Diet)
```

Note that in this particular case, the conversion was not necessary since the `chickweight` table from which we extracted the values stored the `Diet` field as a factor (and this attribute was carried over to our `dat` data frame).

Next, let's plot the points along with the grand mean (dashed line) and each levels' mean value (symbolized as a filled circle).

```{r fig.width=4, fig.height=4}
plot(weight ~ as.numeric(Diet), dat, col=Diet, pch=3,cex=.6)
abline( h = mean(dat$weight), lty=2)
points( unique(dat$Diet) , by(dat$weight, dat[,"Diet"], mean) , pch=19, col=unique(dat$Diet))
```

One level, diet 3, seems to be a bit different from the others, but we will need to run and ANOVA to make sure of this.

```{r}
anova(lm(weight ~ Diet, dat))
```

The small $p$-value suggests that not all levels have the same mean value. We can run a *post hoc* test to identify which level may be different

```{r}
pairwise.t.test( dat$weight, dat$Diet, p.adjust.method = "bonferroni")
TukeyHSD( aov( weight ~ Diet, dat), "Diet")
```

Both tests suggest that diet 3 is significantly different from diet 1, but the differences between the other levels are not significant.


# References
Millard S.P, Neerchal N.K., _Environmental Statistics with S-Plus_, 2001.  
McClave J.T., Dietrich F.H., _Statistics_, 4th edition, 1988.  
Vik P., _Regression, ANOVA, and the General Linear Model: A Statistics Primer_, 2013.
