---
title: "Comparing variances: Fisher's F-test"
editor_options: 
  chunk_output_type: console
---

*Last modified on `r Sys.Date()`*

```{r, echo = FALSE, message = FALSE}
source("libs/Common.R")
```

```{r Load fonts, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
library(extrafont)
font_import(prompt=FALSE, pattern="[A/a]rchitects")
loadfonts(device="postscript")
# loadfonts(device="win") # TO view in windows
```



The test of variances requires that the two sampled population be normally distributed and that the samples are randomly selected from their respective populations.

# Introduction

The method is simple; it consists of taking the ratio between the larger population variance, $\sigma_1^2$, and the smaller population variance, $\sigma_2^2$, then looking up the ratio on an $F$-distribution curve. The null hypothesis states that the ratio equal 1,
$$
H_o: \frac{\sigma_1^2}{\sigma_2^2} = 1
$$
and the alternate hypothesis states that the ratio differs from 1 (i.e. the variances differ),
$$
H_a: \frac{\sigma_1^2}{\sigma_2^2} \neq  1
$$
or is greater than 1 (i.e. $\sigma_1^2$ is significantly bigger than $\sigma_2^2$),
$$
H_a: \frac{\sigma_1^2}{\sigma_2^2} \gt  1
$$

Since the larger variance is assigned to the numerator by convention, we do not have a situation where the ratio is less than 1.

Since we are working with samples, we work with the sample variances $s_1^2$ and $s_2^2$ and compute the test statistic $F$ as follows:
$$
F = \frac{s_1^2}{s_2^2}
$$

The shape of the $\pmb F$**-distribution** curve is defined by both sample's $df$'s, i.e. $(n_1-1)$ and $(n_2-1)$. Like the $\chi^2$ distribution, the $F$ distribution tends to be skewed to the right, especially for large $df$'s.

# Example 1

In one of the examples in the _$z$ and $t$ test_ [tutorial](z_t_tests.html#example-1), we seek to compare the concentration of sulfates between background sites and a contaminated well (data taken from Millard _et al._, p. 418). Did the two samples have equal variances? The table of concentrations is reproduced here.

Contaminated      Background
---------------   -------------------
600               560
590               550
570               570
570               550
565               570
580               590
                  550
                  580

## Solution to example 1

The variances for both samples are $s_{Ref}^2 = 712.5$ and $s_{Cont}^2 = 336.7$. Since $s_{Ref}^2 > s_{Cont}^2$, the value $s_{Ref}^2$ will be in the numerator giving us the following test statistic:

$$
F = \frac{s_{Ref}^2}{s_{Cont}^2} = \frac{712.5}{336.7} = 2.12
$$

Next, we must determine where the $F$ statistic lies along the $F$-distribution curve. This requires that we compute the two $df$'s from the samples to define the shape of the $F$ distribution:
$$
df_{Ref} = 8 - 1 = 7
$$
$$
df_{Cont} = 6 - 1 =5
$$

Now that we have the shape of the $F$-distribution defined, we can look up the probability of getting an $F$ statistic as extreme as ours, An F-distribution table can be used, or the value can be computed exactly using the function `pf()`:

```{r}
pf(2.12, 7, 5, lower.tail=FALSE)
```


```{r echo=FALSE, message=FALSE, fig.width=7, fig.height=3.0, warning=FALSE}
OP <- par("mar"=c(2,4,3,1), xpd=NA)
df1 = 7
df2 = 5
Ft  = 2.12  
xmin = 0
xmax = 10
p10  = qf(0.10,df1,df2)
p025  = qf(0.025,df1,df2)
p975  = qf(0.975,df1,df2)
ncurve.x = seq(xmin, xmax,length.out= 200)
ncurve.y = df(ncurve.x, df1, df2 )
plot( ncurve.y ~ ncurve.x, type="l", ylab=NA, xlab=NA,axes=FALSE,main=NA)
axis(1, family= "Architects Daughter")
lines(x = c(Ft,Ft), y=c(0,df(Ft,df1,df2)), col="red"  , lty=1, lw=2)
lines(x = c(p025,p025) , y=c(0,df( p025,df1,df2)) , col="grey", lty=2, lw=2)
lines(x = c(p975,p975) , y=c(0,df( p975,df1,df2)) , col="grey", lty=2, lw=2)


text(x=Ft, y = df(Ft,df1,df2), eval(Ft), col="red", pos=3, family= "Architects Daughter")
text(x=p025, y = df(p025,df1,df2), "p=0.025", col="#777777", pos=1, family= "Architects Daughter")
text(x=p975, y = df(p975,df1,df2), "p=0.975", col="#777777", pos=3, family= "Architects Daughter")

ncurve.x.mSE <- seq(Ft, xmax, length.out= 100)
ncurve.y.mSE <- df(ncurve.x.mSE, df1, df2 )
ncurve.x.mSE <- c(ncurve.x.mSE, max(ncurve.x.mSE), min(ncurve.x.mSE))
ncurve.y.mSE <- c(ncurve.y.mSE, 0,0)
polygon( ncurve.x.mSE, ncurve.y.mSE, col=rgb(1, 0, 0,.3), border=NA)
par(OP)
```

The $F$ values associated with a probability of 0.025 and 0.975 (associated with rejection regions for a two-tailed $\alpha$ of 0.05) are displayed on the curve in grey dashed vertical lines.

The probability of getting an $F$ as large as ours is about 0.21 (or 21%). Since $H_a$ represents _both_ sides of the distribution, we double the probability to give us the chance of getting a test statistic as great or as small as ours, so for a two-tailed test, $P=0.42$. With such a high $P$-value, we cannot reject the null and therefore can state that for all intents and purposes, the variances between both populations are the same (i.e. the observed variability between both $s$ can be explain by chance alone).

The following figure shows the observed $P$ values in both tails.

```{r echo=FALSE, message=FALSE, fig.width=7, fig.height=3.0, warning=FALSE}
OP <- par("mar"=c(2,4,3,1), xpd=NA)
df1 = 7
df2 = 5
Ft  = 2.12  
xmin = 0
xmax = 10
q.left = .522 # F value assoiatied with 21% probability (in the left tail)
ncurve.x = seq(xmin, xmax,length.out= 200)
ncurve.y = df(ncurve.x, df1, df2 )
plot( ncurve.y ~ ncurve.x, type="l", ylab=NA, xlab=NA,axes=FALSE,main=NA)
axis(1, family= "Architects Daughter")
#lines(x = c(Ft,Ft), y=c(0,df(Ft,df1,df2)), col="red"  , lty=1, lw=2)

#text(x=Ft, y = df(Ft,df1,df2), eval(Ft), col="red", pos=3, family= "Architects Daughter")
text(x=3, y = df(2,df1,df2),  sprintf("Upper 0.21\nregion"), col=rgb(1, 0, 0,.5), pos=4, family= "Architects Daughter")
text(x=0, y = df(.1,df1,df2), sprintf("Lower\n0.21\nregion"), col=rgb(1, 0, 0,.5), pos=2, family= "Architects Daughter")

# Right side
ncurve.x.mSE <- seq(Ft, xmax, length.out= 100)
ncurve.y.mSE <- df(ncurve.x.mSE, df1, df2 )
ncurve.x.mSE <- c(ncurve.x.mSE, max(ncurve.x.mSE), min(ncurve.x.mSE))
ncurve.y.mSE <- c(ncurve.y.mSE, 0,0)
polygon( ncurve.x.mSE, ncurve.y.mSE, col=rgb(1, 0, 0,.3), border=NA)

# Left side
ncurve.x.mSE <- seq(0, q.left, length.out= 100)
ncurve.y.mSE <- df(ncurve.x.mSE, df1, df2 )
ncurve.x.mSE <- c(ncurve.x.mSE, max(ncurve.x.mSE), min(ncurve.x.mSE))
ncurve.y.mSE <- c(ncurve.y.mSE, 0,0)
polygon( ncurve.x.mSE, ncurve.y.mSE, col=rgb(1, 0, 0,.3), border=NA)
par(OP)
```

This can be easily executed in `R` as a two-tailed test as shown in the following code block:

```{r tidy=FALSE}
Ref <-  c(560, 530, 570, 490, 510, 550, 550, 530)
Cont <- c(600, 590, 590, 630, 610, 630)
var.test(Ref, Cont, alternative="two.sided")
```

Note that the `var.test()` computes the $F$ ratio using the first variable name in the list as the numerator. For example, had we reversed the order of variables (i.e. `var.test(Cont, Ref, alternative="two-sided")`), the returned $F$ value would be the inverse  of the original $F$ value, or $1/2.12 = 0.47$. The $P$ value would have stayed the same however.

# Example 2

An investor is concerned that stock 1 is a riskier investment than stock 2 because its variation in daily prices is greater.  The following table provides summary statistics for a sample of 25 daily price changes. 

                     Stock 1   Stock 2
------------------- --------- --------
Sample size          25        25
Standard deviation   .76       .46

Is stock 1's variability significantly greater than that of stock 2, or is the observed difference due to chance?  
_[This example is adapted from McCLave et al. page 461]_

## Solution to example 2

We are asked to test the hypothesis, $H_o$, that the two stock have equal variances and that any observed difference is due to chance (i.e. $\sigma_1^2 = \sigma_2^2$). The alternate hypothesis, $H_a$, states that stock 1 has greater variability than stock 2 (i.e. $\sigma_1^2 > \sigma_2^2$).

Since we are given summary statistics of the samples and not the full dataset, we cannot use the `var.test()` function which requires the full dataset as input. Instead, we will compute the $F$ ratio and observed probabilities separately.

The $F$ ratio is:
$$
F = \frac{(.76)^2}{(.46)^2} = 2.73
$$

The degrees of freedom are $(25 - 1) = 24$ for both samples.

The probability of getting a test statistic as extreme as ours can be computed using the `pf()` function:

```{r tidy=FALSE}
pf( 2.73, 24, 24, lower.tail = FALSE)
```

Note that we are using the  `lower.tail = FALSE` option since our alternate hypothesis is that $\sigma_1^2 > \sigma_2^2$. This gives us a probability of $0.008$, in other words, if the difference between stock 1 and stock 2 were explained by chance variability alone, there would be lest than a 1% chance of computing a $F$ ratio as extreme as ours. We can safely reject $H_o$ and state that the observed difference is real and that stock 1 has greater daily variability than stock 2.

```{r echo=FALSE, message=FALSE, fig.width=7, fig.height=3.0, warning=FALSE}
OP <- par("mar"=c(2,4,3,1), xpd=NA)
df1 = 25
df2 = 25
Ft  = 2.73  
xmin = 0
xmax = 10
p10  = qf(0.10,df1,df2)
p025  = qf(0.025,df1,df2)
p975  = qf(0.975,df1,df2)
ncurve.x = seq(xmin, xmax,length.out= 200)
ncurve.y = df(ncurve.x, df1, df2 )
plot( ncurve.y ~ ncurve.x, type="l", ylab=NA, xlab=NA,axes=FALSE,main=NA)
axis(1, family= "Architects Daughter")
lines(x = c(Ft,Ft), y=c(0,df(Ft,df1,df2)), col="red"  , lty=1, lw=2)

text(x=Ft, y = df(Ft,df1,df2), eval(Ft), col="red", pos=3, family= "Architects Daughter")

ncurve.x.mSE <- seq(Ft, xmax, length.out= 100)
ncurve.y.mSE <- df(ncurve.x.mSE, df1, df2 )
ncurve.x.mSE <- c(ncurve.x.mSE, max(ncurve.x.mSE), min(ncurve.x.mSE))
ncurve.y.mSE <- c(ncurve.y.mSE, 0,0)
polygon( ncurve.x.mSE, ncurve.y.mSE, col=rgb(1, 0, 0,.3), border=NA)
par(OP)
```


# References

Freedman D.A., Robert Pisani, Roger Purves. _Statistics_, 4th edition, 2007.  
McClave J.T., Dietrich F.H., _Statistics_, 4th edition, 1988.  

-----

**Session Info**:

```{r,results='asis', echo=FALSE}
detach("package:extrafont")
pander::pander(sessionInfo(), locale = FALSE)
```


[1]: http://www.amstat.org/sections/srms/pamphlet.pdf
